\chapter{Frequently Asked Questions}
\label{Cha:faq}

\section{SEM2D}
\label{faq_sem2d}

\subsection*{Segmentation fault}

This problem is often related to a small stack size in your computer settings.
In your Linux shell do: \texttt{ulimit -s unlimited} under bash
or \texttt{limit stacksize unlimited} under csh.
Place this command in your startup files (\texttt{.login}, \texttt{.bashrc} or \texttt{.cshrc}).

\subsection*{Instabilities on very distorted elements}

Very distorted elements (with very small or very large angles) 
are usual close to wedges of sedimentary basins, fault branching points, etc.
In general, distorted elements are less stable than square elements:
spurious motions with exponentially increasing amplitude might appear in their vicinity.
In most cases these instabilities can be suppressed by reducing the \texttt{Courant} input parameter.
There is currently no simple recipe to determine the maximum value of this parameter,
so trial and error is required.

\section{EMC2}
\label{faq_emc2}

\subsection*{I can't get rid of a few triangles}

Obtaining a quality quad mesh is not always a trivial task. Trial and error and experience
is needed. This can be by far the most time consuming stage of modeling.

First make sure that the total number of element edges
along the perimeter of each mesh domain is even. This is a necessary
topological condition to generate a quad-only mesh. 

When the geometry seems too complicated for quad meshing you
should consider simplifying the geometry, especially those details that are much smaller
than the dominant wavelength.

If the above fails or does not apply, you have to help the mesher. 
The recommended procedure in EMC2 is:

\begin{enumerate}
\item Divide your original mesh into simple domains,
in such a way that \emph{most} domains have exactly four sides (possibly curved)
and the remaining non-four-sided domains are as small as possible.
\item Generate a structured quad-mesh (a regular grid) inside each four-sided
domain with the \textsf{(QUADRANGULATE)} tool of the \textsf{PREP\_MESH} mode,
as described in section 5.2.13 of EMC2's manual
(note that this is \emph{not} the same as the \textsf{$<$QUADRANGULATE$>$} 
button in the \textsf{EDIT\_MESH} mode). 
\item Proceed as usual (triangulation followed by quadrangulation) 
inside the remaining non-four-sided domains. 
If these are small enough EMC2 should not have problems doing a
correct tri-to-quad meshing.
\end{enumerate}
